\cursedforest\ - A Random Forest Implementation for ``Big'' and ``Wide'' Data \\ Supplementary Information

### 2016-11-26 doc/publications/plos16/plos16.tex
changes

Identifying the relevant or disease specific base-pair difference between individuals is the focus of genome wide
association studies (GWAs).  These analysis are typically performed on only the most frequently differing base-pairs, termed
single nucleotide polymorphisms (SNPs), by applying linear or logistic regression analysis to each SNP separately~\cite{CCC2007}.  It has since been
demonstrated that not taking joint interactions between SNPs into account when filtering for potential disease mutations
is inadequate~\cite{Manolio2009}.  \marginpar{do  these two sentences say teh same thing?}
Complex trait analysis has demonstrated that in order to explain the overall
heritability a model needs to take the joint contribution of all variants into account~\cite{Yang2011}.

@@

Here we extend this toolkit to encompass supervised learning using random forests.  Spark ML's random forest
implementation is not able to handle the extremely ``wide'' genomic data as it was developed for large number of samples
with only modest dimensionality.  Although Spark ML can build a random forest model on a subset of the data (chromosome
1), the time taken is excessive due to the large amount of data being aggregated and processed by the driver node
during intermediate stages of building the model (see Table~\ref{timingtable}). 
\marginpar{missing ref}
 This unbalanced work load where the driver node becomes the bottleneck and worker
nodes being idle prevents a seamless scaling to larger datasets. The memory requirements per executor also increases
with dimensionality due to the data types Spark ML uses, which we elaborate on in the discussion.

@@

The average prediction error for an individual tree $h(X; B)$ is
\marginpar{is this formulation correct in the case of classification? See Friedman on 0-1 losss, Rob}
\begin{equation}
PE_t = E_B E_{X,Y} (Y-h(X; B))^2.
\end{equation}
Assume that, for all $B$, the tree is unbiased, i.e., $EY= E_X h(X; B)$. Then
\begin{equation}
PE_f \leq \rho PE_t
\end{equation}
where $\rho$ is the weighted correlation between residuals $(Y-h(X;B))$ and $(Y-h(X;B^\prime))$ for independent $B,B^\prime$.  
So what is required for accurate random forest regression is (i) low correlation between residuals of differing tree in
the forest, and (ii) low prediction error for the individual trees \cite{Segal.2004}. In addition to growing the trees
on bootstrapped samples, each node in the tree is split on a randomly selected subset of the variables.

>>>>

Identifying the relevant or disease specific base-pair difference between individuals is the focus of genome wide
association studies (GWAs).  These analysis are typically performed on only the most frequently differing base-pairs, termed
single nucleotide polymorphisms (SNPs), by applying linear or logistic regression analysis to each SNP
separately~\cite{CCC2007}.

It has since been
demonstrated that not taking joint interactions between SNPs into account when filtering for potential disease mutations
is inadequate~\cite{Manolio2009,Yang2011}.

@@

Here we extend this toolkit to encompass supervised learning using random forests.  Spark ML's random forest
implementation is not able to handle the extremely ``wide'' genomic data as it was developed for large number of samples
with only modest dimensionality.  Although Spark ML can build a random forest model on a subset of the data (chromosome
1), the time taken is excessive due to the large amount of data being aggregated and processed by the driver node
during intermediate stages of building the model (see Table~\ref{timingtable}). 
This unbalanced work load where the driver node becomes the bottleneck and worker
nodes being idle prevents a seamless scaling to larger datasets. The memory requirements per executor also increases
with dimensionality due to the data types Spark ML uses, which we elaborate on in the discussion.

@@

The average prediction error for an individual tree $h(X; B)$ is

\cite{Friedman.1997}

\begin{equation}
PE_t = E_B E_{X,Y} (Y-h(X; B))^2.
\end{equation}
Assume that, for all $B$, the tree is unbiased, i.e., $EY= E_X h(X; B)$. Then
\begin{equation}
PE_f \leq \rho PE_t
\end{equation}
where $\rho$ is the weighted correlation between residuals $(Y-h(X;B))$ and $(Y-h(X;B^\prime))$ for independent $B,B^\prime$.  
So what is required for accurate random forest regression is (i) low correlation between residuals of differing tree in
the forest, and (ii) low prediction error for the individual trees \cite{Segal.2004}. In addition to growing the trees
on bootstrapped samples, each node in the tree is split on a randomly selected subset of the variables.
